Emergence of the Southeast Asian islands as a driver for Neogene cooling

### 2020-04-08 Manuscript/Supplementary_Information.tex
Updates from Overleaf

\section*{Silicate Weathering Component}

\begin{SCfigure}[0.6][h!]
\begin{center}
	\includegraphics[width=0.6\textwidth]{../Figures/regolith_schematic.pdf}
	\caption{A schematic representation of the silicate weathering component of GEOCLIM in a single profile at steady-state. A rock particle leaves the unweathered bedrock with production rate $P_{r}$, and transits vertically through a regolith of height $h$. Regolith production and physical erosion ($E_{p}$) are equal at steady-state. As the particle transits upwards, some fraction of the primary phases ($x$) are chemically weathered ($W$) with the flux of Ca and Mg corresponding to that weathering rate multiplied by the concentration of Ca+Mg in unweathered bedrock ($\chi_{CaMg}$).}
	\label{fig:regolith_schematic}
\end{center}
\end{SCfigure}

The silicate weathering component of the GEOCLIM model has been modified from the previously published version \citep{Godderis2017b}. The new component implements the model of \citet{Gabet2009a} for the development of a chemically weathered profile. We refer to this chemically weathered profile as regolith where the base of the regolith is unweathered bedrock. In the model of \citet{Gabet2009a}, material enters the regolith and leaves either as a solute through chemical weathering of the material during its travel from the bedrock towards the surface, or as a physically weathered particle once it reaches the top. We use the DynSoil implementation of the \citet{Gabet2009a} model, which integrates chemical weathering within the regolith \citep{West2012a}. The transient time-varying version of this regolith model is described by three equations:

\begin{equation}
    \frac{dh}{dt} = P_{r} - E_{p}
    \label{eq:1}
\end{equation}

\begin{equation}
    \frac{\partial x}{\partial t} = -P_{r} \frac{\partial x}{\partial z} - K \tau^{\sigma}x
    \label{eq:2}
\end{equation}

\begin{equation}
    \frac{\partial \tau}{\partial t} = -P_{r} \frac{\partial \tau}{\partial z} + 1
    \label{eq:3}
\end{equation}

\noindent
Equation \ref{eq:1} is a statement of material conservation, where $h$ is the total height of the regolith (m), $t$ is the model time (yr), $P_{r}$ is the regolith production rate (m/yr), and $E_{p}$ is the physical erosion rate (m/yr). Equation \ref{eq:2} describes how the residual fraction of weatherable phases ($x$, unitless) changes as a function of time ($t$, yr) and depth ($z$; m). $K \tau^{\sigma}$ is the dissolution rate constant, which depends on the local climate (captured by $K$, yr$^{-1-\sigma}$) and the time that a given rock particle has spent in the regolith ($\tau$, yr) to some power $\sigma$ (unitless) which implements a time-dependence. Equation \ref{eq:3} describes how the time that a given rock particle has spent in the regolith changes as time in the model progresses.

The net weathering rate in the regolith column ($W$, m/yr) can then be calculated with:

\begin{equation}
    W = \int_{0}^{h} K \tau^{\sigma} x\;dz
    \label{eq:4}
\end{equation}

The regolith production rate can be expressed as the product of the optimal production rate ($P_{0}$) and a soil production function ($f(h)$):

\begin{equation}
    P_{r} = P_{0}\;f(h)
    \label{eq:5}
\end{equation}

\begin{equation}
    P_{0} = k_{rp}\;q\;e^{-\frac{E_{a}}{R}\left(\frac{1}{T}-\frac{1}{T_{0}}\right)}
    \label{eq:6}
\end{equation}

\begin{equation}
    f(h) = e^{-\frac{-h}{d_{0}}}
    \label{eq:7}
\end{equation}

\noindent
$P_{0}$ is the `optimal' regolith production rate (m/yr), which is defined to be the regolith production rate when there is no overlying regolith. In Equation \ref{eq:6}, we parameterize this value using the model of \citet{Carretier2014a}, where $k_{rp}$ is a proportionality constant (unitless), $q$ is the runoff (m/yr), $E_{a}$ is the activation energy (J/K/mol), $R$ is the ideal gas constant (J/mol), $T$ is the temperature (K), and $T_{0}$ is the reference temperature (K). $f(h)$ is the soil production function (unitless), which describes how regolith production decreases as the thickness of the regolith increases. It takes an exponential form as is commonly applied in the literature (e.g. \citealp{Gabet2009a}). In Equation \ref{eq:7}, we follow \citet{Heimsath1997a}, where $d_{0}$ is a reference regolith thickness (m).

Our implementation of the erosion rate is parameterized based on runoff and slope ($s$; m/m):

\begin{equation}
    E_{p} = k_{e}\;q^{m}\;s^{n}
    \label{eq:8}
\end{equation}

\noindent
$k_{e}$ is a proportionality constant ((m/yr)$^{1-m}$) and $m$ and $n$ are adjustable exponents that are kept as 0.5 and 1 \citep{Maffre2018a}. This formulation is directly inspired by the stream power law \citep{Davy2000a}. This formulation and these exponent values are supported by compilations, but variability in the proportionality constant is difficult to capture at a global scale \citep{Lague2013a}.

The $K$ in the dissolution rate constant in Equation \ref{eq:2} describes the dependence of the chemical weathering on climate:

\begin{equation}
    K = k_{d}\left(1-e^{-k_{w}q}\right)e^{-\frac{E_{a}}{R}\left(\frac{1}{T}-\frac{1}{T_{0}}\right)}
    \label{eq:9}
\end{equation}

\noindent
Equation \ref{eq:9} is an empirical simplification of mineral dissolution rates derived from kinetic theory and laboratory experiments \citep{West2012a}, where $k_{d}$ is a proportionality constant that modifies the dependence of dissolution rate on runoff and temperature (yr$^{-1-\sigma}$), and $k_{w}$ is a proportionality constant that modifies the dependence of dissolution rate on runoff (yr/m).

In this study, we are interested in obtaining the steady-state solution rather than the transient time-varying solution. The steady-state solution for DynSoil can be calculated analytically by setting the time derivatives equal to zero resulting in the following set of equations:

\begin{equation}\label{eq:dynsoil_ss}
\left\{\ 
\begin{aligned}
& h   \ =\    \max\left(  \;  0  \;,\;  d_{0} \log\left(\frac{P_{0}}{E_{p}}\right)  \right)                                                 \\
& x(z)      \ =\   \exp\left(   - \frac{K}{\sigma+1}  \left(\frac{z}{E_{p}}\right)^{\sigma+1}   \right)   \\
& W        \ =\   (1-x(h)) \, E_{p}  \ =\   (1-x_s) \, E_{p}                                                                         \\
\end{aligned}
\right.
\end{equation}

\noindent
$x(z)$ is the abundance profile of primary phases inside the regolith, varying with height upward from the base of the regolith as shown in Figure \ref{fig:regolith_schematic}. Setting $z$ equal to the regolith thickness ($h$) gives $x_s$ which is the proportion of primary phases remaining at the top of the regolith column.

@@

\begin{table}[h!]
\begin{center}
\resizebox{0.7\textwidth}{!}{
    \begin{tabular}{cclcl}
    \textbf{GLiM} & \textbf{GLiM} & \textbf{GLiM} & \textbf{GEOCLIM} & \textbf{GEOCLIM}\\
    \textbf{ID} & \textbf{code} & \textbf{classification} & \textbf{ID} & \textbf{classification}\\
    &&&& \\
    \hline
    &&&& \\
    1 & su & unconsolidated sediments & 6 & sediments \\
    2 & vb & basic volcanic rocks & 4 & mafics \\
    3 & ss & siliciclastic sedimentary rocks & 6 & sediments \\
    4 & pb & basic plutonic rocks & 4 & mafics \\
    5 & sm & mixed sedimentary rocks & 6 & sediments \\
    6 & sc & carbonate sedimentary rocks & 5 & carbonates \\
    7 & va & acid volcanic rocks & 2 & felsics \\
    8 & mt & metamorphics & 1 & metamorphics \\
    9 & pa & acid plutonic rocks & 2 & felsics \\
    10 & vi & intermediate volcanic rocks & 3 & intermediates \\
    11 & wb & water bodies & 0 & water/ice \\
    12 & py & pyroclastics & 2 & felsics \\
    13 & pi & intermediate plutonic rocks & 3 & intermediates \\
    14 & ev & evaporites & 5 & carbonates \\
    15 & nd & no data & 1 & metamorphics \\
    16 & ig & ice and glaciers & 0 & water/ice \\
    \end{tabular}
}
\vspace{10pt}
\caption{Grouping of 16 lithologic categories in GLiM \citep{Hartmann2012a} to 6 broader categories for GEOCLIM.}
\label{tab:GLiM_to_GEOCLIM}
\end{center}
\end{table}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Manuscript/Figures/world_lithology.jpg}
    \caption{Distribution of lithologies at 0.1\degrees $\times$ 0.1\degrees resolution used in GEOCLIM.}
    \label{fig:world_lithology}
\end{figure}

We restricted the calculation of the weathering fluxes to the flux of dissolved Ca+Mg originating from continental silicate weathering, as they are essential for the long-term consumption of atmospheric \COtwo through silicate weathering. To calculate this flux, we need to assign the concentration of Ca and Mg ($\chi_{CaMg}$) within the unweathered bedrock (Fig. \ref{fig:regolith_schematic}). Previous implementations of GEOCLIM have used a ``diffuse lithology'' where all exposed land is assigned the composition of bulk upper continental crust. These previous studies assumed that the weathering rates of all rocks was the same for each continental grid element, provided that the grid element was submitted to the same climatic conditions. Given that the hypothesis we seek to test involves the varying concentration of cations in different lithologies, we instead implement a more realistic lithologically-resolved version of the model.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Manuscript/Figures/W_vs_r2.jpg}
    \caption{\textbf{A)} Modeled global \COtwo consumption vs. the coefficient of determination ($r^{2}$) between modeled and data-constrained \COtwo consumption in each of the watersheds. Each point represents model output using one of the 93,600 parameter combinations (Table \ref{tab:parameter_combinations}). \textbf{B)} Same as A, but zoomed to the plotting space with positive coefficient of determination values. The black line represents the full range of estimates of the present-day non-anthropogenic global CO$_{2}$ emission rate compiled in \citet{Gerlach2011a}. The black box represents the range of these estimates preferred by \citet{Gerlach2011a}.}
    \label{fig:W_vs_r2}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Manuscript/Figures/r2_cross_plot.pdf}
    \caption{Modeled vs. data-constrained \COtwo consumption in watersheds around the world. \textbf{A)} Each point represents a single watershed, and the y-value of the point shows the mean value of the 5,381 parameter combinations that produce individual watershed \COtwo consumption fluxes that approximate those estimated in the literature for the present-day (i.e. the orange points in Figure \ref{fig:W_vs_r2}). Whiskers extend to the minimum and maximum modeled watershed \COtwo consumption fluxes for the 5,381 parameters combinations. The red dashed line is the 1:1 line, the blue dashed line is the 2:1 line, and the cyan dashed lines are the 5:1 and 10:1 lines. \textbf{B)} Same as A, but only showing the parameter combination that produced that highest $r^{2}$. The chemical weathering and regolith thickness maps of this parameter combination are shown in Figure \ref{fig:weathering_regolith_maps}.}
    \label{fig:r2_cross_plot}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Manuscript/Figures/weathering_regolith_maps.jpg}
    \caption{Chemical weathering and regolith thickness maps associated with the GEOCLIM model run using the parameter combination that produced the highest $r^{2}$ (Fig. \ref{fig:r2_cross_plot}B). Parameter values used for this model run are: $k_{d}=5\times10^{-4}$, $k_{w}=1$, $\sigma=-0.4$, $k_{rp}=1\times10^{-2}$, metamorphic$_{Ca+Mg}=2500$~mol/m$^{3}$, sediment$_{Ca+Mg}=2000$~mol/m$^{3}$.}
    \label{fig:weathering_regolith_maps}
\end{figure}

Experimental determinations of the activation energy ($E_a$; Equations \ref{eq:6} and \ref{eq:9}) associated with the weathering of silicate minerals are variable \citep{Brantley2003a}. However, multiple efforts to invert for $E_a$ in basaltic watersheds with varying temperature have yielded values (41.6 $\pm$ 3.2~kJ/mol in \citealp{Li2016a}; 42.3~kJ/mol in \citealp{Dessert2001a}) that are consistent with the lower end of activation energies of Ca+Mg bearing minerals in laboratory experiments such as that for diopside (40.5 $\pm$ 1.7~kJ/mol; \citealp{Knauss1993a}) and for labradorite (42.1~kJ/mol; \citealp{Carroll2005a}). We use the value of 42~kJ/mol in our model runs. While we implement lithology-dependent Ca+Mg concentration, our implementation does not include lithology-dependent kinetics of mineral dissolution. Relative to felsic lithologies, mafic lithologies contain a higher concentration of minerals with faster dissolution kinetics (e.g. plagioclase). However, Ca+Mg from both felsic and mafic lithologies is predominantly sourced from minerals with faster dissolution kinetics, and we therefore use the same chemical weathering formulation (including the same activation energy) across lithologies.

Within the equation that governs the physical erosion rate in GEOCLIM (Equation \ref{eq:8}), the values of the exponents ($m=0.5$ and $n=1$) are supported by compilations \citep{Lague2013a}. However, there is uncertainty in the value of the proportionality constant ($k_e$). We therefore tune this value such that the total erosion flux in the model under present-day slope and runoff conditions (see below) matches the total erosion flux estimated in \citet{Milliman2013a} ($20\times10^{12}$~kg/yr). Assuming that the density of eroded materials is 2500~kg/m$^{3}$, our tuned $k_e$ value is 0.0029110~m$^{1-m}$/yr$^{1-m}$.

Within the equations that govern the chemical weathering rate in GEOCLIM, we identify the less constrained parameters: the proportionality constant that modifies the dependence of dissolution rate on runoff and temperature ($k_{d}$; Equation \ref{eq:9}), the proportionality constant that modifies the dependence of dissolution rate on runoff only ($k_{w}$; Equation \ref{eq:9}), the power constant that modifies the dependence of dissolution rate on the time that a rock particle has spent in the regolith ($\sigma$; Equation \ref{eq:4}), and the proportionality constant that modifies the dependence of regolith production on runoff and temperature ($k_{rp}$; Equation \ref{eq:6}). Furthermore, the Ca+Mg concentrations of metamorphic and siliciclastic sediment grid cells are also difficult to define. We therefore allow these parameters to vary within reasonable bounds during the calibration stage of GEOCLIM. In total, we test 78,000 unique parameter combinations (Fig. \ref{fig:r2_cross_plot}; Table \ref{tab:parameter_combinations}).

@@

We compute spatially-resolved long-term \COtwo consumption (i.e. Ca+Mg fluxes) using present-day runoff (UNH/GRDC Composite Runoff Fields V1.0; \citealp{Fekete1999a}), temperature (CRU TS v.4.03; \citealp{Harris2013a}), and slope (Shuttle Radar Topography Mission; \citealp{Farr2007a}) fields. As described in the main text, we then sum the computed \COtwo consumption over large-scale watersheds that appear in the global compilation of \citet{Gaillardet1999a}, as well as smaller-scale watersheds of the Amazon Basin (HYBAM network) in the compilation of \citet{Moquet2011a, Moquet2016a, Moquet2018a}. The latter are nested watersheds, which requires upstream weathering fluxes to be subtracted from downstream fluxes. Watersheds for which this subtraction yields an aberrant value are not considered. 80 watersheds in total are used in this study. We calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these basins:

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Manuscript/Figures/SEAI_fracs.jpg}
    \caption{0.5\degrees $\times$ 0.5\degrees lithologic maps of the Southeast Asian islands (SEAI) used to force GEOCLIM. Only the 15~Ma scenarios are shown here, but the 10 and 5~Ma scenarios would be similar, using the 10 and 5~Ma shorelines shown in Figure 1 of the main text instead. Each column represents a tested scenario. Each row represents a lithologic category. Solid black lines show present-day shorelines.}
    \label{fig:SEAI_fracs}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Manuscript/Figures/scenario_pCO2_all.pdf}
    \caption{Steady-state \pCOtwo estimates from GEOCLIM for the various scenarios discussed in the text. For each of the 13 scenarios, each point represents an estimate from one of the 5,381 unique parameter combinations that resulted in reasonable total global \COtwo consumption and most closely matched estimates of present-day \COtwo consumption in 80 watersheds around the world (see SI). The box encloses the middle 50\% of the \pCOtwo estimates (i.e. the interquartile range), and the notch represents the median with its 95\% confidence interval. The whiskers extend to the 2.5 and 97.5 percentile values. This figure is identical to Figure 2 in the main text, except for the addition of the scenarios used to test the sensitivity of the results to the inclusion of the Sunda Shelf and the removal of the Greenland ice sheet.}
    \label{fig:scenario_pCO2_all}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.7\textwidth]{Manuscript/Figures/paleogeographic_reconstructions.pdf}
    \caption{Paleogeographically-reconstructed paleoshorelines for SEAI at 5, 10, and 15~Ma, using the paleogeographic model of \citet{Matthews2016a}.}
    \label{fig:paleogeographic_reconstructions}
\end{figure}

@@

The decrease in \pCOtwo that we estimate associated with the emergence of SEAI since 5~Ma (92--161~ppm; Fig. \ref{fig:scenario_pCO2_all}) is greater than that proposed in \citet{Molnar2015a} ($\sim$19~ppm). This 19~ppm value in \citet{Molnar2015a} is obtained from an equation that assumes a direct linear relationship between mean global temperature and changes in weathering-rate-weighted land area, scaled by a factor $\alpha$ that is intended to account for the influence of both runoff and temperature (Equation 3 in \citealp{Molnar2015a}). \pCOtwo is then estimated from $T$ using a simple energy balance equation. However, the relationship between mean global temperature and weathering-rate-weighted land area is not linear. Furthermore, as acknowledged by the authors themselves, this simple linear relationship ignores spatial variability in runoff, temperature, and slope, and only crudely accounts for spatial variability in lithology. In fact, the 19~ppm value estimated by \citet{Molnar2015a} is closer to the decrease in \pCOtwo that we estimate if mafic volcanics associated with the Deccan Traps (a relatively flat area outside of the warm and wet tropics) are replaced with the Ca+Mg concentration of bulk continental crust (21--81~ppm; Fig. \ref{fig:scenario_pCO2_all}). The significant difference in steady-state \pCOtwo estimated between the ``removed Deccan Traps'' scenario and the SEAI scenarios (Fig. \ref{fig:scenario_pCO2_all}) demonstrates that considering changes in the spatial distribution of lithologies alone is not adequate for estimating changes in steady-state \pCOtwo. Instead, spatially-varying runoff, temperature, and slope significantly modulates silicate weathering rates, and must be accounted for.

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Manuscript/Figures/pCO2_proxies.pdf}
    \caption{Proxy-based \pCOtwo estimates for the past 20~m.y. For data that was compiled within \citet{Foster2017a}, error bars indicate standardized uncertainties. For the \citet{Ji2018a} data, error bars represent the 16-84$^{th}$ percentile of resampled estimates. The \citet{Bereiter2015a} data comes from the EPICA Dome C ice core.}
    \label{fig:pCO2_proxies}
\end{figure}

@@

For the paleoshorelines of Sulawesi, we largely followed the recent stratigraphic compilation and paleoshorelines compilation of \citet{Nugraha2018a}. Geographically, Sulawesi can be divided into a central highland region flanked by North, South, Southeast, and East Arms. These arms have high relief ($>$3~km) that are separated by deep basins and broadly define Sulawesi's seven tectonic provinces: 1) the West Sulawesi magmatic arc of the South Arm, 2) the Central Sulawesi metamorphic Belt, 3) the Sangihe arc of the North Arm, 4) the East Sulawesi Ophiolite of the East Arm, 5) the southeast metamorphic belt of the Southeast Arm, and the microcontinental blocks of 6) Banggai-Sula and 7) Buton-Tukang Besi \citep{Hamilton1979a, Katili1978a}. Like SW Borneo, the North and South Arms and much of Central Sulawesi are underlain by continental blocks that rifted off of the Australian-Birds Head margin in the Jurassic and collided in the Cretaceous with Eurasian basement of Sundaland as part of the Woyla Arc system \citep{Parkinson1998b, Hennig2016a, Hall2017a, Hennig2017a}. After collision, subduction polarity reversed and a Cretaceous to Miocene SE-facing volcanic arc developed \citep{Polve1997a, Elburg1998a} that was connected to the paleo-Sunda Arc \citep{Hall2002b}, also referred to as the Great Indonesian arc \citep{Harris2006a}.  The ophiolites of Sulawesi were generated in this arc system in a back arc to intra-arc setting (\citealp{Monnier1995a}, but see \citealp{Kadarusman2004a} for an alternative view), and the largest fragments of the East Sulawesi ophiolite were detached from their metamorphic sole in the late Oligocene (32--28~Ma; \citealp{Parkinson1998a}) and thrust to the east above a west dipping slab (East Sulawesi block of \citealp{Villeneuve2001a}). During the Early to Middle Miocene, the Tukang Besi-Buton block began to collide obliquely with the southeastern end of the Sunda Arc, causing ophiolite emplacement in West and SE Sulawesi, including Buton \citep{Smith1991a, Bergman1996a}.

@@

Paleoshoreline maps were georeferenced from several previous tectonic and paleogeographic analyses, most notably those of \citet{Nichols1991a}, \citet{Cloos2005a}, \citet{Gold2017a}, and \citet{Harrington2017a}. These studies were all based to varying degrees on lithological distributions, biostratigraphic, borehole data, and tectonic models. We complimented these data with our stratigraphic compilations for the region to provide further paleoenvironmental context.

>>>>

@@

\begin{table}[h!]
\begin{center}
\resizebox{0.7\textwidth}{!}{
    \begin{tabular}{cclcl}
    \textbf{GLiM} & \textbf{GLiM} & \textbf{GLiM} & \textbf{GEOCLIM} & \textbf{GEOCLIM}\\
    \textbf{ID} & \textbf{code} & \textbf{classification} & \textbf{ID} & \textbf{classification}\\
    &&&& \\
    \hline
    &&&& \\
    1 & su & unconsolidated sediments & 6 & sediments \\
    2 & vb & basic volcanic rocks & 4 & mafics \\
    3 & ss & siliciclastic sedimentary rocks & 6 & sediments \\
    4 & pb & basic plutonic rocks & 4 & mafics \\
    5 & sm & mixed sedimentary rocks & 6 & sediments \\
    6 & sc & carbonate sedimentary rocks & 5 & carbonates \\
    7 & va & acid volcanic rocks & 2 & felsics \\
    8 & mt & metamorphics & 1 & metamorphics \\
    9 & pa & acid plutonic rocks & 2 & felsics \\
    10 & vi & intermediate volcanic rocks & 3 & intermediates \\
    11 & wb & water bodies & 0 & water/ice \\
    12 & py & pyroclastics & 2 & felsics \\
    13 & pi & intermediate plutonic rocks & 3 & intermediates \\
    14 & ev & evaporites & 5 & carbonates \\
    15 & nd & no data & 1 & metamorphics \\
    16 & ig & ice and glaciers & 0 & water/ice \\
    \end{tabular}
}
\vspace{10pt}
\caption{Grouping of 16 lithologic categories in GLiM \citet{Hartmann2012a} to 6 broader categories for GEOCLIM.}
\label{tab:GLiM_to_GEOCLIM}
\end{center}
\end{table}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/world_lithology.jpg}
    \caption{Distribution of lithologies at 0.1\degrees $\times$ 0.1\degrees resolution used in GEOCLIM.}
    \label{fig:world_lithology}
\end{figure}

We restricted the calculation of the weathering fluxes to the flux of dissolved Ca+Mg originating from continental silicate weathering, as they are essential for the long-term consumption of atmospheric \COtwo through silicate weathering. To calculate this flux, we need to assign the concentration of Ca and Mg ($\chi_{CaMg}$) within the unweathered bedrock (Fig. 1). Previous implementations of GEOCLIM have used a ``diffuse lithology'' where all exposed land is assigned the composition of bulk upper continental crust. These previous studies assumed that the weathering rates of all rocks was the same for each continental grid element, provided that the grid element was submitted to the same climatic conditions. Given that the hypothesis we seek to test involves the varying concentration of cations in different lithologies, we instead implement a more realistic lithologically-resolved version of the model.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/W_vs_r2.jpg}
    \caption{\textbf{A)} Modeled global \COtwo consumption vs. the coefficient of determination ($r^{2}$) between modeled and data-constrained \COtwo consumption in each of the watersheds. Each point represents model output using one of the 93,600 parameter combinations (Table \ref{tab:parameter_combinations}). \textbf{B)} Same as A, but zoomed to the plotting space with positive coefficient of determination values. The black line represents the full range of estimates of the present-day non-anthropogenic global CO$_{2}$ emission rate compiled in \citet{Gerlach2011a}. The black box represents the range of these estimates preferred by \citet{Gerlach2011a}.}
    \label{fig:W_vs_r2}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/r2_cross_plot.pdf}
    \caption{Modeled vs. data-constrained \COtwo consumption in watersheds around the world. \textbf{A)} Each point represents a single watershed, and the y-value of the point shows the mean value of the 5,381 parameter combinations that produce individual watershed \COtwo consumption fluxes that approximate those estimated in the literature for the present-day (i.e. the orange points in Figure \ref{fig:W_vs_r2}). Whiskers extend to the minimum and maximum modeled watershed \COtwo consumption fluxes for the 5,381 parameters combinations. The red dashed line is the 1:1 line, the blue dashed line is the 2:1 line, and the cyan dashed lines are the 5:1 and 10:1 lines. \textbf{B)} Same as A, but only showing the parameter combination that produced that highest $r^{2}$. The chemical weathering and regolith thickness maps of this parameter combination are shown in Figure \ref{fig:weathering_regolith_maps}.}
    \label{fig:r2_cross_plot}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/weathering_regolith_maps.jpg}
    \caption{Chemical weathering and regolith thickness maps associated with the GEOCLIM model run using the parameter combination that produced the highest $r^{2}$ (Fig. \ref{fig:r2_cross_plot}B). Parameter values used for this model run are: $k_{d}=5\times10^{-4}$, $k_{w}=1$, $\sigma=-0.4$, $k_{rp}=1\times10^{-2}$, metamorphic$_{Ca+Mg}=2500$~mol/m$^{3}$, sediment$_{Ca+Mg}=2000$~mol/m$^{3}$.}
    \label{fig:weathering_regolith_maps}
\end{figure}

Experimental determinations of the activation energy ($E_a$; Equations 6 and 9) associated with the weathering of silicate minerals are variable \citep{Brantley2003a}. However, multiple efforts to invert for $E_a$ in basaltic watersheds with varying temperature have yielded values (41.6 $\pm$ 3.2~kJ/mol in \citealp{Li2016a}; 42.3~kJ/mol in \citealp{Dessert2001a}) that are consistent with the lower end of activation energies of Ca+Mg bearing minerals in laboratory experiments such as that for diopside (40.5 $\pm$ 1.7~kJ/mol; \citealp{Knauss1993a}) and for labradorite (42.1~kJ/mol; \citealp{Carroll2005a}). We use the value of 42~kJ/mol in our model runs. While we implement lithology-dependent Ca+Mg concentration, our implementation does not include lithology-dependent kinetics of mineral dissolution. Relative to felsic lithologies, mafic lithologies contain a higher concentration of minerals with faster dissolution kinetics (e.g. plagioclase). However, Ca+Mg from both felsic and mafic lithologies is predominantly sourced from minerals with faster dissolution kinetics, and we therefore use the same chemical weathering formulation (including the same activation energy) across lithologies.

Within the equation that governs the physical erosion rate in GEOCLIM (Equation 8), the values of the exponents ($m=0.5$ and $n=1$) are supported by compilations \citep{Lague2013a}. However, there is uncertainty in the value of the proportionality constant ($k_e$). We therefore tune this value such that the total erosion flux in the model under present-day slope and runoff conditions (see below) matches the total erosion flux estimated in \citet{Milliman2013a} ($20\times10^{12}$~kg/yr). Assuming that the density of eroded materials is 2500~kg/m$^{3}$, our tuned $k_e$ value is 0.0029110~m$^{1-m}$/yr$^{1-m}$.

Within the equations that govern the chemical weathering rate in GEOCLIM, we identify the less constrained parameters: the proportionality constant that modifies the dependence of dissolution rate on runoff and temperature ($k_{d}$; Equation 9), the proportionality constant that modifies the dependence of dissolution rate on runoff only ($k_{w}$; Equation 9), the power constant that modifies the dependence of dissolution rate on the time that a rock particle has spent in the regolith ($\sigma$; Equation 4), and the proportionality constant that modifies the dependence of regolith production on runoff and temperature ($k_{rp}$; Equation 6). Furthermore, the Ca+Mg concentrations of metamorphic and siliciclastic sediment grid cells are also difficult to define. We therefore allow these parameters to vary within reasonable bounds during the calibration stage of GEOCLIM. In total, we test 78,000 unique parameter combinations (Fig. \ref{fig:r2_cross_plot}; Table \ref{tab:parameter_combinations}).

@@

We compute spatially-resolved long-term \COtwo consumption (i.e. Ca+Mg fluxes) using present-day runoff (UNH/GRDC Composite Runoff Fields V1.0; \citealp{Fekete1999a}), temperature (CRU TS v.4.03; \citealp{Harris2013a}), and slope (Shuttle Radar Topography Mission; \citealp{Farr2007a}) fields. As described in the main text, we then sum the computed \COtwo consumption over large-scale watersheds that appear in the global compilation of \citet{Gaillardet1999a}, as well as smaller-scale watersheds of the Amazon Basin (HYBAM network) in the compilation of \citet{Moquet2011a}, \citet{Moquet2016a}, and \citet{Moquet2018a}. The latter are nested watersheds, which requires upstream weathering fluxes to be subtracted from downstream fluxes. Watersheds for which this subtraction yields an aberrant value are not considered. 80 watersheds in total are used in this study. We calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these basins:

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/SEAI_fracs.jpg}
    \caption{0.5\degrees $\times$ 0.5\degrees lithologic maps of the Southeast Asian islands (SEAI) used to force GEOCLIM. Only the 15~Ma scenarios are shown here, but the 10 and 5~Ma scenarios would be similar, using the 10 and 5~Ma shorelines shown in Figure 1 of the main text instead. Each column represents a tested scenario. Each row represents a lithologic category. Solid black lines show present-day shorelines. Total areas of each lithologic category within the SEAI for each of these tested scenarios is shown in Figure \ref{fig:lithology_areas}.}
    \label{fig:SEAI_fracs}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Manuscript/Figures/lithology_areas.pdf}
    \caption{Total area of the lithologic categories within the SEAI for each tested scenario. Note that the grey bars in the middle panel indicate the fractional area of each lithologic category for the entire Earth.}
    \label{fig:lithology_areas}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/scenario_pCO2_all.pdf}
    \caption{Steady-state \pCOtwo estimates from GEOCLIM for the various scenarios discussed in the text. For each of the 13 scenarios, each point represents an estimate from one of the 5,381 unique parameter combinations that resulted in reasonable total global \COtwo consumption and most closely matched estimates of present-day \COtwo consumption in 80 watersheds around the world (see SI). The box encloses the middle 50\% of the \pCOtwo estimates (i.e. the interquartile range), and the notch represents the median with its 95\% confidence interval. The whiskers extend to the 2.5 and 97.5 percentile values. This figure is identical to Figure 2 in the main text, except for the addition of the scenarios used to test the sensitivity of the results to the inclusion of the Sunda Shelf and the removal of the Greenland ice sheet.}
    \label{fig:scenario_pCO2_all}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.7\textwidth]{Figures/paleogeographic_reconstructions.pdf}
    \caption{Paleogeographically-reconstructed paleoshorelines for SEAI at 5, 10, and 15~Ma, using the paleogeographic model of \citet{Matthews2016a}.}
    \label{fig:paleogeographic_reconstructions}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/pCO2_proxies.pdf}
    \caption{Proxy-based \pCOtwo estimates for the past 20~m.y. For data that was compiled within \citet{Foster2017a}, error bars indicate standardized uncertainties. For the \citet{Ji2018a} data, error bars represent the 16-84$^{th}$ percentile of resampled estimates. The \citet{Bereiter2015a} data comes from the EPICA Dome C ice core.}
    \label{fig:pCO2_proxies}
\end{figure}

@@

For the paleoshorelines of Sulawesi, we largely followed the recent stratigraphic compilation and paleoshorelines compilation of \citet{Nugraha2018a}. Geographically, Sulawesi can be divided into a central highland region flanked by North, South, Southeast, and East Arms. These arms have high relief ($>$3~km) that are separated by deep basins and broadly define Sulawesi's seven tectonic provinces: 1) the West Sulawesi magmatic arc of the South Arm, 2) the Central Sulawesi metamorphic Belt, 3) the Sangihe arc of the North Arm, 4) the East Sulawesi Ophiolite of the East Arm, 5) the southeast metamorphic belt of the Southeast Arm, and the microcontinental blocks of 6) Banggai-Sula and 7) Buton-Tukang Besi \citep{Hamilton1979a, Katili1978a}. Like SW Borneo, the North and South Arms and much of Central Sulawesi are underlain by continental blocks that rifted off of the Australian-Birds Head margin in the Jurassic and collided in the Cretaceous with Eurasian basement of Sundaland as part of the Woyla Arc system \citep{Parkinson1998b, Hennig2016a, Hall2017a, Hennig2017a}. After collision, subduction polarity reversed and a Cretaceous to Miocene SE-facing volcanic arc developed \citep{Polve1997a, Elburg1998a} that was connected to the paleo-Sunda Arc \citep{Hall2002b}, also referred to as the Great Indonesian arc \citep{Harris2006a}.  The ophiolites of Sulawesi were generated in this arc system in a back arc to intra-arc setting \citep{Monnier1995a} (but see \citealp{Kadarusman2004a} for an alternative view) and the largest fragments of the East Sulawesi ophiolite were detached from their metamorphic sole in the late Oligocene (32--28~Ma; \citealp{Parkinson1998a}) and thrust to the east above a west dipping slab (East Sulawesi block; \citealp{Villeneuve2001a}). During the Early to Middle Miocene, the Tukang Besi-Buton block began to collide obliquely with the southeastern end of the Sunda Arc, causing ophiolite emplacement in West and SE Sulawesi, including Buton \citep{Smith1991a, Bergman1996a}.

@@

Paleoshoreline maps were georeferenced from several previous tectonic and paleogeographic analyses, most notably those of \citet{Nichols1991a, Cloos2005a, Gold2017a, Harrington2017a}. These studies were all based to varying degrees on lithological distributions, biostratigraphic, borehole data, and tectonic models. We complimented these data with our stratigraphic compilations for the region to provide further paleoenvironmental context.
